---
title: "WT and Gata5/6 MO 13hpf combined data Seurat analysis"
author: "Xuefei Yuan"
date: "11/15/2020"
output: html_document
---

```{r "setup", include=FALSE}
knitr::opts_chunk$set(echo = TRUE, fig.width = 4, fig.height = 5)
knitr::opts_knit$set(root.dir = '/home/xuefei/Github/')
```

### Summary
This file contains analyses needed for making Fig.2 and Supplementary Fig.2, starting from the gene-cell count matrix


### Preliminaries {-}
```{r, message=FALSE}
### packages and functions needed
library(Seurat)
library(ggplot2)
library(sctransform)
source("customized_function_seurat.R")
```


### Data loading and filtering
```{r data_loading, message=FALSE, cache=T, warning=FALSE}
## ## We load the gene-cell count matrix (output from CellRanger) to creat a Seurat Object and merge the WT and MO datasets

gbm <- merge(
  x = CreateSeuratObject(counts = Read10X("matrix_data/wt_13hpf_gbm/"),min.cells = 3,  project = "wt_13hpf"), 
  y = CreateSeuratObject(counts = Read10X("matrix_data/mo_13hpf_gbm/"),min.cells = 3, project = "g56mo_13hpf"), 
  add.cell.id1 = "wt.13hpf", add.cell.id2 = "mo.13hpf", project = "13hpf_2sample")

### calculate the percentage of mitochondrial reads in each cell 
gbm <- PercentageFeatureSet(object = gbm, pattern = "^mt-", col.name = "percent.mt")

```

### Plot the distribution of UMI counts, number of genes detected in the raw data
```{r quality_check, message=FALSE, warning=FALSE}
## fuction for plotting the distribution of nCount_RNA, nFeature_RNA, percent.mt figure
plot_nGene_nUMI_mito(gbm = gbm, raw_data=T) 
```

### Filter low-quality cells and potential doublets then plot the distribution again
```{r filter, cache=T}
gbm <- subset(x = gbm, subset = nCount_RNA < 30000 & nFeature_RNA > 1000 & percent.mt < 4)
## fuction for plotting the distribution of nCount_RNA, nFeature_RNA, percent.mt figure
plot_nGene_nUMI_mito(gbm = gbm, raw_data=F) 
```



### SCTransform for normalization, find variable genes, data scaling and PCA
```{r SCTransform, eval=FALSE}
### cell cycle genes determined in PMID: 26430063
cc.genes <- readLines(con ="./other_files/regev_lab_cell_cycle_genes_zebrafish.txt")
s.genes <- cc.genes[1:43]
g2m.genes <- cc.genes[44:96]
gbm <- CellCycleScoring(object = gbm, s.features = s.genes, g2m.features = g2m.genes, set.ident = F)

gbm <- SCTransform(object = gbm, vars.to.regress = c("percent.mt", "S.Score", "G2M.Score", "nCount_RNA"), verbose = T, variable.features.n = 3000) 

gbm <- RunPCA(object = gbm, verbose = T, npcs = 60)
gbm <- JackStraw(object = gbm, num.replicate = 100, dims = 50)
gbm <- ScoreJackStraw(object = gbm, dims = 1:50)

```

```{r echo=FALSE}
## some colde above and below for demonstration purposes, load the preprocessed data to save time
load("./Rdata/wt_g56KD_13hpf.RData")
```


### Jack Straw and Elbow plot to determine the PC numbers to use
```{r JackStrawPlot,warning=F, cache=T, fig.width=9, fig.height=5}
JackStrawPlot(object = gbm, dims = 1:30)

```

```{r ElbowPlot, warning=F, cache=T, fig.width=4, fig.height=4}
ElbowPlot(object = gbm, ndims = 60)
```

### UMAP visualization and clustering
```{r UMAP, eval=FALSE}
gbm <- RunUMAP(object = gbm, dims = 1:20, verbose = T, min.dist = 0.3,
               n.neighbors = 30, n.epochs = 500, set.op.mix.ratio = 0.5)

gbm <- FindNeighbors(object = gbm, dims = 1:20, verbose = T, prune.SNN = 1/15, k.param = 20)
gbm <- FindClusters(object = gbm, verbose = T, resolution = 1.5)
## note that in Seurat cluster numbers start with 0, but we change it to starting with 1 in all our figures
```

```{r umap_clusters, warning=FALSE, fig.width = 6, fig.height = 5}
## plot by cluster
DimPlot(object = gbm, label = TRUE, reduction = "umap", pt.size =0.5)
```

```{r umap_sample, warning=FALSE, fig.width = 7, fig.height = 5}
## plot by sample
DimPlot(object = gbm, label = TRUE, pt.size =0.5, group.by = "orig.ident" )
```

### Differential analysis to identify marker genes for each cluster
```{r find_markers,eval=FALSE}
### run all different methods
method.ls <- list("wilcox", "bimod", "roc", "poisson", "negbinom", "LR", "MAST")

diff.glob.ls <- lapply(method.ls, 
                       function(x) FindAllMarkers(gbm, min.pct = 0.1, logfc.threshold = 0.25, 
                                                  only.pos = T, test.use=x, return.thresh = 0.01))
names(diff.glob.ls) <- method.ls
### plot the marker genes identified by different methods and save in the heatmaps as pdf
lapply(method.ls, function(x) plotHeatmap(gbm=gbm, test.use = x, topN=40))

```

### select the top genes to plot
```{r plot_heatmap_1, warning=FALSE, fig.width = 9, fig.height = 7, message=FALSE}
## genes plotted in supplementary Figure
genes.plot <- c("thbs3b","cthrc1a", "twist1b","fsta","kirrelb",
                "meis1a",  "aldh1a2", "meis2b", "dhrs3a", "FRMD1", 
                "foxj1a", "hnf1bb", "acy3.2","pax2a","gem",
                "gata1a",  "cldng", "gfi1aa","tal1","morc3b",
                "col15a1b", "marcksl1b","pknox2", "foxf2a","pitx3",
                "foxa2", "mcama","prdx5", "cd9b", "bcam", 
                "hand2", "cdx4","pdgfra", "msx3","nos1apa",
                "hoxa11b", "hoxd10a", "msx2b","tbx16","bambia",
                "eve1", "msgn1",  "cyp26a1", "snai1a","her7",
                "mef2ca", "ttn.2", "nkx2.5", "bmp6","mef2cb",
                "foxa3","onecut1", "nr5a2", "limk2", "hbbe1.2", ## liver endoderm
                "hbbe1.1", "hbae3", "myb","blf", "klf3",
                "irx1b", "eya2", "rgl1", "rspo3","efna5a",
                "fli1b", "she", "kdrl","cdh5","hey2",
                 "he1b",  "ctslb",
                 "hyou1", "inhbaa",
                "myod1", "ttn.1", 
                "cldni",  "cyt1" ## epidermis
                
                )
### genes plotted in the main figure
genes.plot2 <- c("thbs3b","cthrc1a",
                 "meis2b", "dhrs3a",
                 "foxj1a", "hnf1bb",
                 "gata1a", "gfi1aa",
                 "col15a1b", "pknox2",
                 "foxa2", "cd9b", 
                 "hand2", "pdgfra",
                 "hoxa11b", "msx2b",
                 "eve1", "msgn1",
                 "mef2ca", "ttn.2",
                 "onecut1", "nr5a2", ## liver endoderm
                 "hbbe1.1", "myb", 
                 "irx1b", "eya2",
                 "fli1b", "kdrl",
                 "ctslb","he1b",
                 "hyou1", "inhbaa",
                 "myod1", "ttn.1", 
                 "cldni",  "cyt1" ## epidermis
                 
)

### make the dot plot in the main figure
DotPlot.m(object = gbm, features = genes.plot2, dot.scale = 4,split.by = "orig.ident")

```

```{r plot_heatmap_2, warning=FALSE, fig.width = 9, fig.height = 12}
### heatmap in supplementary figures
DoHeatmap(object = gbm, features = genes.plot)
```

```{r plot_heatmap_3, warning=FALSE, fig.width = 18, fig.height = 12}
DoHeatmap.m(object = gbm, features = genes.plot, split.by = "orig.ident")
```

### Calculate cell composition changes
```{r cell_comp_cal}
perc_table <- prop.table(x = table(gbm@active.ident, gbm@meta.data[, "orig.ident"]), 
                         margin = 2)
#table(gbm.2@ident, gbm.2@meta.data$orig.ident)
perc.df <- data.frame("wt.13hpf"=perc_table[, 2],
                      "gata56MO.13hpf"=perc_table[, 1])
perc.df$FC <- sapply(rownames(perc.df), function(x) (perc.df[x, "gata56MO.13hpf"]-perc.df[x, "wt.13hpf"])/perc.df[x, "wt.13hpf"])
perc.df <- perc.df[order(perc.df$FC, decreasing = F), ]
perc.df$cluster_id <- rownames(perc.df)
### make sure sort the plot by FC
perc.df$cluster_id <- factor(perc.df$cluster_id, levels = perc.df$cluster_id)
### To use the same color as that used in TSNE cluster plot
perc.df$color_col <- c("09", "17", "07", "06", "10", "02",
                       "11", "08", "05", "13", "15", "14", "04", 
                       "16", "01", "03" , "00", "12")

### calculate if the proportion change is significant
prop.df <- getStats_FC(gbm=gbm) ### stat results table

perc.df <- cbind(perc.df, prop.df[rownames(perc.df) , 2:3])
perc.df$stars <- "*"
perc.df[perc.df$p.adj.Fisher > 0.05, "stars"] <- NA

```

```{r cell_comp_plot}
### bar plot
ggplot(perc.df)+
  geom_col(aes(cluster_id, log10(FC+1), fill=color_col))+
  guides(fill=F)+coord_flip()+
  ylab("Log10(% in gata56MO/% in WT)")


## dot plot
ggplot(perc.df)+
  geom_point(aes(cluster_id, log10(FC+1), color=color_col, size=wt.13hpf))+
  guides(color=F)+coord_flip()+scale_y_continuous(limits = c(-2, 0.5))+
  scale_size_continuous(breaks = c("1%"=0.01, "4%"=0.04, "7%"=0.07, "10%"=0.10))+
  ylab("Log10(% in gata56MO/% in WT)")

```

### Calculate cell composition changes after the cardiac cluster is removed
```{r cell_comp_cal2}
### remove the cardiac cluster and caculate the fold change again to see if other expansions were driven by this cardiac depletion
noCardi.stat.df <- getStats_FC(gbm = gbm, toRemove = c("9"))

noCardi_table <- prop.table(x = table(gbm@active.ident, gbm@meta.data[, "orig.ident"])[-c(10), ], margin = 2)
#table(gbm.2@ident, gbm.2@meta.data$orig.ident)
noCardi.df <- data.frame("wt.13hpf"=noCardi_table[, 2],
                         "gata56MO.13hpf"=noCardi_table[, 1])
noCardi.df$FC <- sapply(rownames(noCardi.df), 
                        function(x) (noCardi.df[x, "gata56MO.13hpf"]-noCardi.df[x, "wt.13hpf"])/noCardi.df[x, "wt.13hpf"])
### make sure sort the plot by FC
noCardi.df <- noCardi.df[order(noCardi.df$FC, decreasing = F), ]
noCardi.df$cluster_id <- rownames(noCardi.df)

### To make the color scheme consistent with the original plot, mannually add a row to the table with artificial value
### set FC to NA so it won't be ploted at the end
noCardi.df <- rbind(data.frame(wt.13hpf=0.05,
                               gata56MO.13hpf=0.001,
                               FC=NA,
                               cluster_id="9"),
                    noCardi.df)
rownames(noCardi.df) <- noCardi.df$cluster_id

noCardi.df$cluster_id <- factor(noCardi.df$cluster_id, levels = noCardi.df$cluster_id)

### To use the same color as that used in UMAP cluster plot
noCardi.df$color_col <- c("09","17", "07", "06", "10", "02",
                       "11", "08", "05", "13", "15", "14", "04", 
                       "16", "01", "03" , "00", "12")
noCardi.df <- cbind(noCardi.df, noCardi.stat.df[rownames(noCardi.df) , 2:3])
noCardi.df$stars <- "*"
noCardi.df$p.adj.Fisher[1] <- 1
noCardi.df[noCardi.df$p.adj.Fisher > 0.05, "stars"] <- NA
#noCardi.df$p.Fisher <- NULL
#noCardi.df$p.adj.Fisher <- NULL
#### use the percentage in real WT samples for plotting
noCardi.df$wt.13hpf <-perc.df$wt.13hpf

```

```{r cell_comp_plot2, warning=F, message=F}
### bar plot
ggplot(noCardi.df)+
  geom_col(aes(cluster_id, log10(FC+1), fill=color_col))+
  guides(fill=F)+coord_flip()+
  ylab("Log10(% in gata56MO/% in WT)")

### dot plot
ggplot(noCardi.df)+
  geom_point(aes(cluster_id, log10(FC+1), color=color_col, size=wt.13hpf))+
  guides(color=F)+coord_flip()+
  scale_y_continuous(limits = c(-1, 0.5))+
  scale_size_continuous(breaks = c("1%"=0.01, "4%"=0.04, "7%"=0.07, "10%"=0.10))+
  ylab("Log10(% in gata56MO/% in WT)")

```


### Save Seurat object for subsetting analyses
```{r eval=F}
saveRDS(gbm, file = "./Rdata/wt_g56KD_13hpf.rds")
```

