---
title: "WT 8hpf data Seurat analysis"
author: "Xuefei Yuan"
date: "11/15/2020"
output: html_document
---

```{r "setup", include=FALSE}
knitr::opts_chunk$set(echo = TRUE, fig.width = 4, fig.height = 5)
knitr::opts_knit$set(root.dir = '/home/xuefei/Github/')
```

### Summary
This file contains analyses needed for Supplementary Fig.1, starting from the gene-cell count matrix


### Preliminaries {-}
```{r, message=FALSE}
### packages and functions needed
library(Seurat)
library(ggplot2)
library(sctransform)
source("customized_function_seurat.R")
```


### Data loading and filtering
```{r data_loading, message=FALSE, cache=T, warning=FALSE}
## ## We load the gene-cell count matrix (output from CellRanger) to creat a Seurat Object.

gbm <- CreateSeuratObject(counts = Read10X("./matrix_data/wt_8hpf_gbm/"),
                          min.cells = 3, project = "wt_8hpf")

### calculate the percentage of mitochondrial reads in each cell 
gbm <- PercentageFeatureSet(object = gbm, pattern = "^mt-", col.name = "percent.mt")

```

### Plot the distribution of UMI counts, number of genes detected in the raw data
```{r quality_check, message=FALSE, warning=FALSE}
## fuction for plotting the distribution of nCount_RNA, nFeature_RNA, percent.mt figure
plot_nGene_nUMI_mito(gbm = gbm, raw_data=T) 
```

### Filter low-quality cells and potential doublets then plot the distribution again
```{r filter, cache=T, message=FALSE, warning=FALSE}
gbm <- subset(x = gbm, subset = nCount_RNA < 20000 & nFeature_RNA > 1000 & percent.mt < 4)
## fuction for plotting the distribution of nCount_RNA, nFeature_RNA, percent.mt figure
plot_nGene_nUMI_mito(gbm = gbm, raw_data=F) 
```



### SCTransform for normalization, find variable genes, data scaling and PCA
```{r SCTransform, eval=FALSE}
### cell cycle genes determined in PMID: 26430063
cc.genes <- readLines(con = "./other_files/regev_lab_cell_cycle_genes_zebrafish.txt")
s.genes <- cc.genes[1:43]
g2m.genes <- cc.genes[44:96]
gbm <- CellCycleScoring(object = gbm, s.features = s.genes, g2m.features = g2m.genes, set.ident = F)

gbm <- SCTransform(object = gbm, vars.to.regress = c("percent.mt", "S.Score", "G2M.Score", "nCount_RNA"), 
                   verbose = T, variable.features.n = 3000) 

gbm <- RunPCA(object = gbm, verbose = T, npcs = 60)
gbm <- JackStraw(object = gbm, num.replicate = 100, dims = 50)
gbm <- ScoreJackStraw(object = gbm, dims = 1:50)

```

```{r echo=FALSE}
## some colde above and below for demonstration purposes, load the preprocessed data to save time
load("./Rdata/wt_8hpf.RData")
```


### Jack Straw and Elbow plot to determine the PC numbers to use
```{r JackStrawPlot,warning=F, cache=T, fig.width=7, fig.height=4.5}
JackStrawPlot(object = gbm, dims = 1:30)

```

```{r ElbowPlot, warning=F, cache=T, fig.width=4, fig.height=3.5}
ElbowPlot(object = gbm, ndims = 60)
```

### UMAP visualization and clustering
```{r UMAP, eval=FALSE}
gbm <- RunUMAP(object = gbm, dims = 1:20, verbose = T, min.dist = 0.2,
               n.neighbors = 30, n.epochs = 500, set.op.mix.ratio = 0.5)

gbm <- FindNeighbors(object = gbm, dims = 1:20, verbose = T, prune.SNN = 1/15, k.param = 20)
gbm <- FindClusters(object = gbm, verbose = T, resolution = 1.5)
## note that in Seurat cluster numbers start with 0, but we change it to starting with 1 in all our figures
```

```{r clusters, warning=FALSE, fig.width = 6, fig.height = 5}
## plot by cluster
DimPlot(object = gbm, label = TRUE, reduction = "umap", pt.size =0.5)
```


### Differential analysis to identify marker genes for each cluster
```{r find_markers,eval=FALSE}
### run all different methods
method.ls <- list("wilcox", "bimod", "roc", "poisson", "negbinom", "LR", "MAST")

diff.glob.ls <- lapply(method.ls, 
                       function(x) FindAllMarkers(gbm, min.pct = 0.1, logfc.threshold = 0.25, 
                                                  only.pos = T, test.use=x, return.thresh = 0.01))
names(diff.glob.ls) <- method.ls
### plot the marker genes identified by different methods and save in the heatmaps as pdf for examination
lapply(method.ls, function(x) plotHeatmap(gbm=gbm, test.use = x, topN=40, diff.ls = diff.glob.ls))

```

### select the top genes to plot
```{r plot_heatmap_1, warning=FALSE, fig.width = 9, fig.height = 7, message=FALSE}
## genes plotted in supplementary Figure
genes.plot <- c("sox32", "sox17", "apela", "sp5l", "ca9", "vox", 
                 "rdh10a","dlx3b", "meis3" ,"hand2", "gata6", "foxh1",
                 "hes6", "tbx16", "rbm38", "tbx6l", "msgn1", "lbx2", 
                 "prdx5", "her5", "sox13", "abtb2a", "tfa", "txn", 
                 "cxcl12b", "twist1a", "lratb", "tbx1", "cyp26c1", "irx1b",
                 "fsta", "ripply1",  "gbx1", "meis2b", "nr2f6b", "vcana",
                  "tram1","ctslb", "calr3b", "ssr4", "ctsc", "ctslb.1",
                 "xbp1", "cited4b", "frzb", "fzd8b","btg2","gsc",
                 "msx3", "szl", "smad7", "pdgfra", "id1", "tead1b",
                 "ccng1", "phlda3", "NUPR2", "mdm2", "sesn3", "tp53",
                 "alcamb","pcdh19", "mdkb", "tfap2a", "sox19a", "sox3")

```

```{r plot_heatmap_2, warning=FALSE, fig.width = 8, fig.height = 9.5}
### heatmap in supplementary figures
DoHeatmap(object = gbm, features = genes.plot)
```



### Save Seurat object for other analyses
```{r eval=F}
saveRDS(gbm, file = "./Rdata/wt_8hpf.rds")
```

